---
title: "Code Appendix"
author: "Jaylin Lowe and Jenna Korobova"
date: "6/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library(HLMdiag)
library(wesanderson)
library(gridExtra)
```

# Data Manipulation 

```{r Data Cleanup}
# original dataset
housing <- read_csv("housingAll.csv")

# filter to just have apartments, prices less than 5000 (reducing skew and variabiltiy)
# removing region (not very useful variable) and electric_vehicle_charge (too few observations)
housing2 <- housing %>%
  filter(type == "apartment", price <= 5000) %>%
  select(-c(region, electric_vehicle_charge))

n <- nrow(housing2)

# choosing random sample of 2000 observations from original dataset to be our new dataset
set.seed(790)
index <- sample(n, 2000, replace = FALSE)
housing_subset <- housing2[index,]

# making new csv
write.csv(housing_subset, "housingSubset.csv")

# dataframe
data <- read_csv("housingSubset.csv")
```

# EDA

```{r Tables}
# checking that all apartments
table(data$type) 

# tables of counts of level one variables
table(data$cats_allowed)
table(data$dogs_allowed)
table(data$smoking_allowed)
table(data$wheelchair_access)
table(data$comes_furnished)
```

```{r Distribution of Price}
# idistribution of price is skewed right
ggplot(data) + geom_histogram(aes(x = price), color = "white", bins = 25)

# price distribution is more symmetric when removed prices of 0 and 1
# and log-transformed price
data %>%
  filter(price > 5) %>%
  ggplot() + geom_histogram(aes(x = log(price)), color = "white", bins = 25)
```

```{r Price by State}
# median
data %>%
  group_by(`State Name`) %>%
  summarize(MedianPrice = median(price)) %>%
  ggplot() + geom_bar(aes(x = reorder(`State Name`, MedianPrice), y = MedianPrice), stat = "identity") + coord_flip()

# mean
data %>%
  group_by(`State Name`) %>%
  summarize(MeanPrice = mean(price)) %>%
  ggplot() + geom_bar(aes(x = reorder(`State Name`, MeanPrice), y = MeanPrice), stat = "identity") + coord_flip()
``` 

Top 5 states: DC, Hawaii, New Jersey, New Hampshire, California
Bottom 5 states: Kansas, Missouri, Oklahoma, West Virginia, Mississippi 

```{r}
summary(data2$price)
sd(data2$price)

summary(log(data2$price))
sd(log(data2$price))

table(data2$Governor)
table(data2$pets_allowed)
table(data2$smoking_allowed)

summary(data2$beds)
sd(data2$beds)

summary(data2$baths)
sd(data2$baths)

summary(data2$PopDensity)
sd(data2$PopDensity)

summary(data2$TaxRate)
sd(data2$TaxRate)

summary(data2$MedianIncome)
sd(data2$MedianIncome)

summary(data2$sqfeet)
sd(data2$sqfeet)
```

```{r Boxplots of Price by Level One Dummy Variables}
# as number of bedrooms increases, price tends to increase
ggplot(data, aes(x = beds, y = price)) + geom_point() + geom_jitter()

# making level one dummy variables factor variables for plotting
data$cats_allowed2 <- factor(data$cats_allowed, levels = c(0,1))
data$dogs_allowed2 <- factor(data$dogs_allowed, levels = c(0,1))
data$smoking_allowed2 <- factor(data$smoking_allowed, levels = c(0,1))
data$wheelchair_access2 <- factor(data$wheelchair_access, levels = c(0,1))
data$comes_furnished2 <- factor(data$comes_furnished, levels = c(0,1))

# grid of boxplots of price by respective level one dummy variable
p1 <- ggplot(data) + 
  geom_boxplot(mapping = aes(y = price, x = cats_allowed2)) + 
  labs(x = "Cats Allowed") 
  
p2 <- ggplot(data) + 
  geom_boxplot(mapping = aes(y = price, x = dogs_allowed2)) + 
  labs(x = "Dogs Allowed") 

p3 <- ggplot(data) + 
  geom_boxplot(mapping = aes(y = price, x = smoking_allowed2)) + 
  labs(x = "Smoking Allowed") 

p4 <- ggplot(data) + 
  geom_boxplot(mapping = aes(y = price, x = wheelchair_access2)) + 
  labs(x = "Wheelchair Access") 

p5 <- ggplot(data) + 
  geom_boxplot(mapping = aes(y = price, x = comes_furnished2)) + 
  labs(x = "Comes Furnished")  

grid.arrange(p1, p2, p3, p4, p5, ncol = 2)

# wheelchair_access, comes_furnished, and smoking_allowed have slightly different
# distibutions depending on dummy variable value—we should look out for those
```

```{r Boxplots of Price by Level Two Dummy Variables}
#scale_fill_manual(values = wes_palette("Moonrise2", n = 2)) 

data.new <- na.omit(data) # DC has no governor
# making DeathPenalty a factor dummy variable
data.new$DeathPenalty2 <- factor(data.new$DeathPenalty, levels = c(0,1)) 

p1.2 <- ggplot(data.new) + 
  geom_boxplot(mapping = aes(y = price, x = Governor)) + 
  labs(x = "Governor Political Party") 

p2.2 <- ggplot(data.new) + 
  geom_boxplot(mapping = aes(y = price, x = DeathPenalty2)) + 
  labs(x = "Death Penalty in State") 

grid.arrange(p1.2, p2.2, ncol = 1) 

# looks like states without the death penalty and democratic governors
# tend to have higher priced rent
```

```{r Rent Prices by State}
# median price
data %>%
  group_by(`State Name`) %>%
  summarize(MedianPrice = median(price)) %>%
  ggplot() + geom_bar(aes(x = reorder(`State Name`, MedianPrice), y = MedianPrice), stat = "identity") + coord_flip()

# mean price
data %>%
  group_by(`State Name`) %>%
  summarize(MeanPrice = mean(price)) %>%
  ggplot() + geom_bar(aes(x = reorder(`State Name`, MeanPrice), y = MeanPrice), stat = "identity") + coord_flip()
``` 

Top 5 states: DC, Hawaii, New Jersey, New Hampshire, California
Bottom 5 states: Kansas, Missouri, Oklahoma, West Virginia, Mississippi 

```{r Level One and Two Variables}
#scale_fill_manual(values = wes_palette("Moonrise2", n = 2)) 

data.new2 <- subset(data, subset = sqfeet < 3500)

# linetype isn't an ideal aesthetic here but it helps a little
# as sqfeet increases, price increases
ggplot(data.new2, mapping = aes(y = price, x = sqfeet, group = as.factor(beds))) + 
  geom_point(aes(color = as.factor(beds))) +
  geom_smooth(aes(linetype = as.factor(beds)), method = "lm", se = FALSE, color = "black")
  labs(title = "Scatterplot of Sqft by Price") 
  
# no strong relationship between PctBlack and price it seems
ggplot(data, mapping = aes(y = price, x = PctBlack)) +
  geom_point()
 
data[which(data$PopDensity > 7500),c(15, 18:20)] # population density in DC is ridiculous 

# DC has incredibly high popdensity, might lead to some outliers, let's try subsetting it out
ggplot(data, mapping = aes(y = price, x = PopDensity)) +
  geom_point()

# as popdensity increases, price increases, likely because those are cities
data.new3 <- subset(data, subset = state != "dc") %>%
  group_by(state)

# as PopDensity increases, price tends to increase too, likely because cities have
# higher PopDensity and therefore also higher prices
ggplot(data.new3, mapping = aes(y = price, x = PopDensity)) +
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm")

# TaxRate doesn't seem to be very indicative of price
ggplot(housing_subset, mapping = aes(y = price, x = TaxRate)) +
  geom_point() + 
  geom_smooth(se = FALSE, method = "lm")
```

# Modeling

```{r Subsetted Final Data Based on EDA}
# getting rid of prices of 0 and 1, DC, and collapsing cats_allowed and dogs_allowed into one variable
data2 <- data %>%
  filter(price > 5) %>%
  filter(state != "dc") %>%
  mutate(pets_allowed = ifelse(cats_allowed == 1 & dogs_allowed == 1, 1, 0))
```

## Random Effects

```{r Random Effects, cache = TRUE}
# after Laura's suggestions

# just random intercept
housing.lmer4.1 <- lmer(log(price) ~ scale(sqfeet) + beds + baths + (1|state), data = data2)
summary(housing.lmer4.1)

# random slope
housing.lmer4.2 <- lmer(log(price) ~ scale(sqfeet) + beds + baths + (scale(sqfeet) + beds + baths|state), data = data2)
summary(housing.lmer4.2) # beds and baths are very correlated (-0.93)

# alt - null
D <- 2*(logLik(housing.lmer4.2)-logLik(housing.lmer4.1))
N <- 500
Dsim <- numeric(N)
nullY <- simulate(housing.lmer4.1, nsim = N)
for (i in 1:N) {
  null.lmer <- refit(housing.lmer4.1, nullY[,i])
  alt.lmer <- refit(housing.lmer4.2, nullY[,i])
  Dsim[i] <- 2*(logLik(alt.lmer) - logLik(null.lmer))
}
hist(Dsim)
abline(v = D)
mean(Dsim >= D) # p-value is 1, just need random slopes
```

## Fixed Effects

```{r Fixed Effects}
# all fixed effects and random slope
housing.lmer.fix1 <- lmer(log(price) ~ scale(sqfeet) + beds + baths + pets_allowed + smoking_allowed + wheelchair_access + comes_furnished + scale(MedianIncome) + Governor + PopDensity + TaxRate + DeathPenalty + PctWhite + PctBlack + (1|state), data = data2, REML = FALSE)
summary(housing.lmer.fix1)

# remove wheelchair_acess, DeathPenalty, and comes_furnished from housing.lmer.fix1
housing.lmer.fix2 <- lmer(log(price) ~ scale(sqfeet) + beds + baths + pets_allowed + smoking_allowed + scale(MedianIncome) + Governor + PopDensity + TaxRate + PctWhite + PctBlack + (1|state), data = data2, REML = FALSE)
summary(housing.lmer.fix2)

# alt - null
D <- 2*(logLik(housing.lmer.fix1)-logLik(housing.lmer.fix2))
1-pchisq(D, 3) # p-val is 0.9324294, prefer smaller model, housing.lmer.fix2

# removing Pctlack, pctWhite, TaxRate from housing.lmer.fix2
housing.lmer.fix3 <- lmer(log(price) ~ scale(sqfeet) + beds + baths + pets_allowed + smoking_allowed + scale(MedianIncome) + Governor + PopDensity + (1|state), data = data2, REML = FALSE)
summary(housing.lmer.fix3)

# alt - null
D <- 2*(logLik(housing.lmer.fix2)-logLik(housing.lmer.fix3))
1-pchisq(D, 3) # p-val is 0.02657733, prefer larger model, housing.lmer.fix2

# removing just Pctlack, pctWhite from housing.lmer.fix2
housing.lmer.fix4 <- lmer(log(price) ~ scale(sqfeet) + beds + baths + pets_allowed + smoking_allowed + scale(MedianIncome) + Governor + PopDensity + TaxRate + (1|state), data = data2, REML = FALSE)
summary(housing.lmer.fix4)

# alt - null
D <- 2*(logLik(housing.lmer.fix2)-logLik(housing.lmer.fix4))
1-pchisq(D, 2) # p-val is 0.086, prefer smaller model, housing.lmer.fix4

# no other low t-values, done checking fixed effects
```

## Interactions

```{r Interaction Terms}
# working model is housing.lmer.fix4

# maybe beds and baths interact?
housing.lmer.int1 <- lmer(log(price) ~ scale(sqfeet) + beds*baths + pets_allowed + smoking_allowed + scale(MedianIncome) + Governor + PopDensity + TaxRate + (1|state), data = data2, REML = FALSE)
summary(housing.lmer.int1)

# alt - null
D <- 2*(logLik(housing.lmer.int1)-logLik(housing.lmer.fix4))
1-pchisq(D, 1) # p-val is 0, prefer housing.lmer.fix4, no interactions
```

## Working Model

```{r Final Model Maybe}
housing.lmer.fix4 <- lmer(log(price) ~ scale(sqfeet) + beds + baths + pets_allowed + smoking_allowed + scale(MedianIncome) + Governor + PopDensity + TaxRate + (1|state), data = data2, REML = FALSE)
summary(housing.lmer.fix4)
```

# Diagnostics

```{r Diagnostics One}
x <- HLMresid(housing.lmer.fix4, level = "marginal", standardize = TRUE)

#standardized marginal: curvature? outliers?
plot(x ~ sqfeet, data = data2)
abline(h= 0) # three bad ones
#residuals in order of observation number
plot(x)
abline(h = 0) # three bad ones
#standardized conditional

x2 <- HLMresid(housing.lmer.fix4 , level = 1, standardize = TRUE)
#outliers?
plot(x2) #in order of observation number
abline(h= 0)

which(x < -6) # observations 79, 1579, and 1770 need to be removed
```

```{r Refit}
# removing the three points from the dataset
data3 <- data2[-c(79, 1579, 1770), ]

# refitting housing.lmer.fix4 without those three observations
housing.lmer.fix4.2 <- lmer(log(price) ~ scale(sqfeet) + beds + baths + pets_allowed + smoking_allowed + scale(MedianIncome) + Governor + PopDensity + TaxRate + (1|state), data = data3, REML = FALSE)
summary(housing.lmer.fix4.2)
```

```{r Diagnostics Two}
xNew <- HLMresid(housing.lmer.fix4.2, level = "marginal", standardize = TRUE)

#standardized marginal: curvature? outliers?
plot(xNew ~ sqfeet, data = data3)
abline(h= 0) # huge cluster
#residuals in order of observation number
plot(xNew)
abline(h = 0) # much better
#standardized conditional

x2New <- HLMresid(housing.lmer.fix4.2, level = 1, standardize = TRUE)
#outliers?
plot(x2New) #in order of observation number
abline(h= 0) # looks okay

cd <- cooks.distance(housing.lmer.fix4.2)
plot(cd, type = "h") # one super high cook's D, let's remove it

which(cd > 0.03) # removing 1229 and 1231 will make the rest of the points okay
```

```{r Refit Two}

# removing the three points from the dataset
data3 <- data3[-c(1229, 1231), ]

# refitting housing.lmer.fix4 without those two observations
housing.lmer.fix4.3 <- lmer(log(price) ~ scale(sqfeet) + beds + baths + pets_allowed + smoking_allowed + scale(MedianIncome) + Governor + PopDensity + TaxRate + (1|state), data = data3, REML = FALSE)
summary(housing.lmer.fix4.3)
```

```{r Diagnostics Final}
cd2 <- cooks.distance(housing.lmer.fix4.3)
plot(cd2, type = "h") 

lev <- leverage(housing.lmer.fix4.3, level = 1)
head(lev)
```

# Final Model

Refit using new dataset, `data3`, with observations 79, 1229, 1231, 1579, and 1770 removed.

```{r}
housing.lmer.fix4.3 <- lmer(log(price) ~ scale(sqfeet) + beds + baths + pets_allowed + smoking_allowed + scale(MedianIncome) + Governor + PopDensity + TaxRate + (1|state), data = data3, REML = FALSE)
summary(housing.lmer.fix4.3)
```
